Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-08 doc/thesis/intro.tex
Fix final spelling and grammar mistakes

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor its descendant, or the difference in average
body mass or height. A phylogeny with branch lengths in calendar time units is
often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually correspond to the present day, as a reference
point~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
\cref{fig:speciestree}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population~\autocite{ypma2013relating}. Like
phylogenies, transmission trees have tips, nodes, edges, and branch lengths.
However, rather than recording an evolutionary process (speciation), they
record an epidemiological process (transmission). The tips of a transmission
tree represent the removal of infected hosts, while internal nodes correspond
to transmissions from one host to another. Transmission trees generally have
branch lengths in units of calendar time, with branching times indicating times
of transmission. The root of a transmission tree corresponds to the initially
infected patient who introduced the epidemic into the network, also known as
the \defn{index case}. The internal nodes may be labelled with the donor of the
transmission pair, if this is known. The tips of the tree, rather than being
fixed at the present day, are placed at the time at which the individual was
removed from the epidemic, such as by death, recovery, isolation, behaviour
change, or migration~\autocite{stadler2013uncovering}. Consequently, the
transmission tree may not be ultrametric, but may have tips located at varying
distances from the root. Such trees are said to have \defn{heterochronous}
taxa~\autocite{drummond2003measurably}, in contrast to the \defn{isochronous}
taxa found in most phylogenies of macro-organisms. A transmission tree is
illustrated in \cref{fig:contactnet} (right).

Each infected individual in an epidemic may appear in the transmission tree
more than once. This is different from the transmission \emph{network}, in
which each infected individual appears exactly once, and edges are in
one-to-one correspondence with transmissions~\autocite{welch2011statistical,
keeling2005networks}. The distinction between the two objects is illustrated in
\cref{fig:contactnet}. However, since transmission networks generally have no
cycles (unless re-infection occurs), they are trees in the graph theoretical
sense, and hence are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}. The term
``transmission network'' is taken to mean the subgraph of the contact network
along which transmissions occurred, following
\eg~\autocite{welch2011statistical, keeling2005networks}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its descendants. The \defn{production rule} of
$n$ is its total number of children, and its number of leaf children. That is,
if two nodes have the same number of children and among these, the same number
of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
Gaussian radial basis function of the vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidian norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\[
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum{n_2 \in N(T_2)} \Delta (n_1, n_2),
\]
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.

@@

\Gls{SIS} is one type of \gls{SMC} method, whose aim is to sample from a
distribution $\pi$ on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1,
\ldots, x_d)$. The basis of \gls{SIS} is \gls{IS}, which is a method of
estimating summary statistics of distributions which are known only up to a
normalizing constant, and therefore cannot be sampled from directly. That is,
if $\pi$ is such a distribution and $f$ is any real-valued function, \gls{IS}
is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x') = \int \eta_{i-1}(x) K_i(x, x') \d x,
\end{align}
where $K_i$ is a Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(x, x') = \max\left(1, \frac{q(x', x)\pi_i(x)}{q(x, x')\pi_i(x')}\right),
\]
where $q(x, x')$ is a proposal function such as a Gaussian distribution
centered at $x$ (see \cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractible. In
particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} Markov kernels
$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary, and will be
precisely defined on a problem-specific basis. Second, they define an
alternative sequence of target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
of increasing dimension. This brings us back to the setting described above in
\cref{subsec:sis}, namely of building up an importance distribution of
dimension $d$ sequentially through lower-dimensional distributions. We can
write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
\begin{align*}
  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
\end{align*}
and hence
\[
  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
\]
Therefore, the importance weights for these new targets are defined recursively as
\begin{align}
  w(\vec{x_i}) 
    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &= w(\vec{x_{i-1}}) \cdot
      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &\propto w(\vec{x_{i-1}}) \cdot
      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
    \label{eq:smcwt}
\end{align}
The final key piece of information is to notice that, because the $L_i$ are
Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
These are all the ingredients we need to apply \gls{SIS}. The sequences of
kernels $L$ and $K$ should be chosen based on the problem at hand to inimize
the variance in the importance weights as well as possible. For a fixed choice
of $K_i$, the backward kernels $L_i$ which minimize this variance are called
the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
presented as \cref{alg:smcsamp}. A resampling step is applied whenever the
\gls{ESS} of the population drops too low, as discussed in the previous
section.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The intuition for why this approximation might hold comes from the fact that, 
when $\varepsilon = 0$, the intergral on the left is exactly equal to the
posterior. Thus, by taking $\varepsilon$ small, we should attain something
close to the posterior.

The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the  \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. Note that in many formulations, the distance function $\rho$
is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which maps
data points into a vector of summary statistics. This can be useful if the data
are high-dimensional or of a complex type, but it is not strictly necessary.
For instance, if the data are numeric and of low dimension, the distance
function may simply be the Euclidian distance~\autocite{sisson2007sequential}.
For more complex data, \textcite{nakagome2013kernel} proposed the use of a
kernel function (defined in \cref{subsec:treeshape}), an approach they dubbed
\defn{kernel-\gls{ABC}}.

@@

All the algorithms discussed in this section can be straightforwardly extended
to sample from the joint distribution
\[
  \pi_\varepsilon(\theta, z_1, \ldots, z_M \mid y),
\]
which is equivalent to associating $M$ simulated datasets to each parameter
particle instead of just one. The simulated dataset $z$ is replaced by
$z = z_1, \ldots, z_M$, and the indicator function for the
$\varepsilon$-ball around $y$ is replaced by
\[
  \sum_{k=1}^M \I_{A_{\varepsilon, y}} (z_i).
\]
For \gls{ABC}-\gls{MCMC} and \gls{ABC}-\gls{SMC}, the proposal distribution
$q(\theta, \theta') f(z \mid \theta')$ is replaced by
\[
  q_i(\theta, \theta') \prod_{k=1}^M f(z_i \mid \theta').
\]
%\subsection{Convergence properties}
%
%The convergence of \gls{ABC} to the true posterior distribution of interest
%depends on two components: the convergence of the underlying algorithm to the
%\gls{ABC} target distribution, and the resemblance of this target distribution
%to the true posterior~\autocite{marin2012approximate}. We shall address these 
%components one at a time. Rejection \gls{ABC} and \gls{ABC}-\gls{MCMC} use,
%respectively, rejection sampling and \gls{MCMC} to sample from the target
%distributions. Both of these methods are well-established and their convergence
%properties have been extensively studied. Since we do not apply rejection or
%\gls{MCMC} in this work, we shall restrict our attention to the convergence
%properties of \gls{SMC}.

%Many results about the rate of convergence of \gls{MCMC} are available, as well
%as convergence diagnostics. \Gls{SMC} is a much newer method than \gls{MCMC}\ldots

>>>>

The branches of the tree may have associated lengths, representing either
evolutionary distance or calendar time between ancestors and their descendants.
The term ``evolutionary distance'' is used here imprecisely to mean any sort of
quantitative measure of evolution, such as the number of differences between
the DNA sequences of an ancestor and its descendant, or the difference in
average body mass or height. A phylogeny with branch lengths in calendar time
units is often referred to as \defn{time-scaled}. In a time-scaled phylogeny,
the internal nodes can be mapped onto a timeline by using the tips of the tree,
which usually correspond to the present day, reference
points~\autocite{nee1992tempo}. The corresponding points on the timeline are
called \defn{branching times}, and the rate of their accumulation is referred
to as the \defn{branching rate}. Rooted trees whose tips are all the same
distance from the root are called \defn{ultrametric}
trees~\autocite{buneman1974note}. These concepts are illustrated in
\cref{fig:speciestree}.

@@

In epidemiology, a \defn{transmission tree} is a graphical representation of an
epidemic's progress through a population~\autocite{ypma2013relating}. Like
phylogenies, transmission trees have tips, nodes, edges, and branch lengths.
However, rather than recording an evolutionary process (speciation), they
record an epidemiological process (transmission). The tips of a transmission
tree represent the removal by sampling of infected hosts, while internal nodes
correspond to transmissions from one host to another. Transmission trees
generally have branch lengths in units of calendar time, with branching times
indicating times of transmission. The root of a transmission tree corresponds
to the initially infected patient who introduced the epidemic into the network,
also known as the \defn{index case}. The internal nodes may be labelled with
the donor of the transmission pair, if this is known. The tips of the tree,
rather than being fixed at the present day, are placed at the time at which the
individual was removed from the epidemic, such as by death, recovery,
isolation, behaviour change, or migration~\autocite{stadler2013uncovering}.
Consequently, the transmission tree may not be ultrametric, but may have tips
located at varying distances from the root. Such trees are said to have
\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
transmission tree is illustrated in \cref{fig:contactnet} (right). 
{\color{blue} The object on the right of the figure is called a \defn{contact
network}, which depicts the entire susceptible population along with all
possible routes of disease transmission. Contact networks, and their
relationships to transmission trees, will be discussed further in
\cref{sec:contactnet}}.

Each infected individual in an epidemic may appear at nodes of the transmission
tree more than once. This is different from the transmission \emph{network}, in
which each infected individual appears exactly once, and edges are in
one-to-one correspondence with transmissions~\autocite{welch2011statistical,
keeling2005networks}. The distinction between the two objects is illustrated in
\cref{fig:contactnet}. However, since transmission networks generally have no
cycles (unless re-infection occurs), they are trees in the graph theoretical
sense, and hence are sometimes also referred to as transmission
trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
``transmission tree'' for the objects depicted on the right side of
\cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}. The term
``transmission network'' is taken to mean the subgraph of the contact network
along which transmissions occurred, following
\eg~\autocite{welch2011statistical, keeling2005networks}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its descendants. The \defn{production rule} of
$n$ is its total number of children, and its number of leaf children. That is,
if two nodes have the same number of children and among these, the same number
of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
Gaussian radial basis function of the vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\[
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum{n_2 \in N(T_2)} \Delta (n_1, n_2),
\]
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.

@@

\Gls{SIS}~\autocite{gordon1993novel} is one type of \gls{SMC} method, whose aim
is to sample from a distribution $\pi$ on an high-dimensional space, say
$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
which is a method of estimating summary statistics of distributions which are
known only up to a normalizing constant, and therefore cannot be sampled from
directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
function, \gls{IS} is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

@@

Both algorithms involve progression through a sequence of related
distributions. For \gls{SIS}, these distributions are lower-dimensional
marginals of the target distribution, while for the \gls{SMC} sampler, they are
of the same dimension and constitute a smooth progression from an initial to a
final distribution. In both cases, the neighbouring distributions in the
sequence are related to each other in some way, and we can take advantage of
that relationship to create a sequence of importance distributions alongside
the sequence of targets. In \gls{SIS}, the neighbouring marginals
$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
distribution. In \gls{SMC}, the relationship between subsequent distributions
is less explicit, but it is assumed that they are related closely enough that
an importance distribution for $\pi_i$ can be easily transformed into one for
$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
is constructed as
\begin{align}
  \label{eq:impint}
  \eta_i(x') = \int \eta_{i-1}(x) K_i(x, x') \d x,
\end{align}
where $K_i$ is a Markov kernel and the integral is over the space on which the
$\pi_i$ are defined. The choice of $K_i$ should be based on the perceived
relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
That is,
\[
  K_i(x, x') = \max\left(1, \frac{q(x', x)\pi_i(x)}{q(x, x')\pi_i(x')}\right),
\]
where $q(x, x')$ is a proposal function such as a Gaussian distribution
centred at $x$ (see \cref{subsec:mfit}).

Although this method of building up $\eta$ appears straightforward, the
drawback is that the importance distribution itself becomes intractable. In
particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
perform \gls{IS}, this construction appears to have defeated the purpose of
providing an importance distribution for each $\pi_i$.
\textcite{del2006sequential} overcome this problem with two ``artificial''
objects. First, they propose the existence of \textit{backward} Markov kernels
$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary, and will be
precisely defined on a problem-specific basis. Second, they define an
alternative sequence of target distributions
\[
  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
\]
of increasing dimension. This brings us back to the setting described above in
\cref{subsec:sis}, namely of building up an importance distribution of
dimension $d$ sequentially through lower-dimensional distributions. We can
write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
\begin{align*}
  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
\end{align*}
and hence
\[
  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
\]
Therefore, the importance weights for these new targets are defined recursively as
\begin{align}
  w(\vec{x_i}) 
    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &= w(\vec{x_{i-1}}) \cdot
      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
    &\propto w(\vec{x_{i-1}}) \cdot
      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
    \label{eq:smcwt}
\end{align}
The final key piece of information is to notice that, because the $L_i$ are
Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
These are all the ingredients we need to apply \gls{SIS}. The sequences of
kernels $L$ and $K$ should be chosen based on the problem at hand to minimize
the variance in the importance weights as well as possible. For a fixed choice
of $K_i$, the backward kernels $L_i$ which minimize this variance are called
the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
presented as \cref{alg:smcsamp}. A resampling step is applied whenever the
\gls{ESS} of the population drops too low, as discussed in the previous
section.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The intuition for why this approximation might hold comes from the fact that, 
when $\varepsilon = 0$, the integral on the left is exactly equal to the
posterior. Thus, by taking $\varepsilon$ small, we should attain something
close to the posterior.

The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the  \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. Note that in many formulations, the distance function $\rho$
is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which maps
data points into a vector of summary statistics. This can be useful if the data
are high-dimensional or of a complex type, but it is not strictly necessary.
For instance, if the data are numeric and of low dimension, the distance
function may simply be the Euclidean distance~\autocite{sisson2007sequential}.
For more complex data, \textcite{nakagome2013kernel} proposed the use of a
kernel function (defined in \cref{subsec:treeshape}), an approach they dubbed
\defn{kernel-\gls{ABC}}.

@@

All the algorithms discussed in this section can be straightforwardly extended
to sample from the joint distribution
\[
  \pi_\varepsilon(\theta, z_1, \ldots, z_M \mid y),
\]
which is equivalent to associating $M$ simulated datasets to each parameter
particle instead of just one. The simulated dataset $z$ is replaced by
$z = z_1, \ldots, z_M$, and the indicator function for the
$\varepsilon$-ball around $y$ is replaced by
\[
  \sum_{k=1}^M \I_{A_{\varepsilon, y}} (z_i).
\]
For \gls{ABC}-\gls{MCMC} and \gls{ABC}-\gls{SMC}, the proposal distribution
$q(\theta, \theta') f(z \mid \theta')$ is replaced by
\[
  q_i(\theta, \theta') \prod_{k=1}^M f(z_i \mid \theta').
\]

%\subsection{Convergence properties}
%
%The convergence of \gls{ABC} to the true posterior distribution of interest
%depends on two components: the convergence of the underlying algorithm to the
%\gls{ABC} target distribution, and the resemblance of this target distribution
%to the true
%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
%address these components one at a time. Rejection \gls{ABC} and
%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
%sample from the target distributions. Both of these methods are
%well-established and their convergence properties have been extensively
%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
%restrict our attention to the convergence properties of \gls{SMC}.
